---
title: "benthos-trends-dataprep"
author: "Tom Webb"
date: "2023-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Temporal trends from presence-absence data in European benthos

This document desrcibes the workflow for assembling the data sets required to produce spatial estimates of temporal turnover (species loss and gain, beta diversity) from the EMODnet macrobenthos presence-absence product.

First, load required packages:

```{r, load_packages, message = FALSE}
library(tidyverse)
library(ncdf4)
library(here)
library(worrms)
```

The main presence-absence dataset is provided as a NetCDF file. This will be served via EMODnet but is accessed locally here. First, establish a connection to the file:

```{r, open_nc_connection}
nc_fil <- nc_open(here("data", "EMODnet-Biology-Benthos-European-Seas/Macrobenthos_Eur_Seas_Pres_Abs_v0-4.nc"))
```

### Assembling a data set of unique sampling events

Assemble a data frame of sampling events, composed of latitude, longitude, full date, and year from the .nc file:

```{r, get_sampling_events}
sample_events <- tibble(
  lat = as.vector(ncvar_get(nc_fil, "lat")),
  lon = as.vector(ncvar_get(nc_fil,"lon")),
  date = lubridate::ymd(as.Date(ncvar_get(nc_fil, "Date"), origin = "1970-01-01 00:00:00")),
  year = lubridate::year(as.Date(ncvar_get(nc_fil, "Date"), origin = "1970-01-01 00:00:00"))
)
```

Note that there are fewer distinct sampling events (`r n_distinct(sample_events)`) than there are total sampling events (`r nrow(sample_events`). Note also that a few sampling events have dates in the future:
```{r, future_samples, echo = FALSE}
sample_events %>% filter(date > lubridate::now())
```

### Assembling a dataset of unique species IDs

First get all taxon names and WoRMS Aphia IDs from the nc file:
```{r, get_taxa}
taxon_name <- ncvar_get(nc_fil, "Taxon_Name")
aphia_id <- ncvar_get(nc_fil,"AphiaID") %>% word(start = -1, sep = ":") %>% as.integer()
```

For this product, we use only taxa at the species rank only. To do this this we need to get the WoRMS classification for each taxon, and take the final entry in this. For this we use a simple function:

```{r, get_taxonrank}
get_worms_rank <- function(aphiaid){
  classif <- wm_classification(aphiaid) %>% slice(n())
  classif
}
```

To run for one taxon:
```{r, get_taxonrank_example}
get_worms_rank(aphia_id[1000])
```
This will run the function over all taxa. *WARNING*: this takes ~40 minutes on my machine:

```{r, get_taxonrank_alltaxa, eval = FALSE}
taxo <- aphia_id %>% map_df(get_worms_rank, .progress = TRUE)
```

Because this is slow to run, the output was written to file (`write_csv(taxo, file = here("data", "EMODnet-Biology-Benthos-European-Seas/benthos_taxo_rank.csv"))`), so it can simply be read in directly here:

```{r, read_taxo, message = FALSE}
taxo <- read_csv(here("data", "EMODnet-Biology-Benthos-European-Seas/benthos_taxo_rank.csv"))
```

What ranks are present?
```{r, print_taxonranks}
taxo %>% count(rank) %>% arrange(desc(n)) %>% print(n = 100)
```

To get a list of only species-rank taxa (adding an index for ease of matching later):
```{r, get_speciesrank}
spp <- taxo %>%
  filter(rank == "Species") %>% 
  mutate(aphia_index = match(.$AphiaID, taxo$AphiaID))
```

This results in `r nrow(spp)` unique species-level benthic taxa.

### Get occurrences of a species by sampling event

We can get presence-absence for a single species directly from the nc file:

```{r, get_presabs_singlespecies}
presabs <- ncvar_get(nc_fil,"Pres_abs",
                     start = c(1, spp$aphia_index[1]), count = c(-1,1))
```

To run this across species, wrap it in a function:

```{r, get_presabs_species}
get_presabs_sp <- function(sp_index, sp_df = spp, samps = sample_events){
  
  # get pres-abs data
  presabs <- ncvar_get(nc_fil, "Pres_abs",
                       start = c(1, sp_index), count = c(-1,1))
  # assemble data frame
  ret <- tibble(aphia_id = spp$AphiaID[spp$aphia_index == sp_index],
                presabs = presabs)
  ret <- samps %>% bind_cols(ret)
  
  ret
}

```

Run this for a single species and get a count of presences and absences:

```{r, get_presabs_eg}

presabs <- get_presabs_sp(sp_index = spp$aphia_index[1])
presabs %>% count(presabs)

```

NB: 1 = Present, 0 = looked for but absent, NA = not looked for. A value of 'NA' indicates that at that sampling event, only a subset of the full benthos community was effectively surveyed. To work out how big an issue this is, we can get species-level summaries of numbers of presences, absences, and NAs, in a function:

```{r, get_presabs_summary}
get_presabs_sp_summ <- function(sp_index, sp_df = spp){
  
  # get pres-abs data
  presabs <- ncvar_get(nc_fil, "Pres_abs",
                       start = c(1, sp_index), count = c(-1,1))
  # assemble data frame
    ret <- tibble(presabs = presabs) %>%
      count(presabs) %>%
      mutate(aphia_id = spp$AphiaID[spp$aphia_index == sp_index]) %>% 
      dplyr::select(aphia_id, everything())
  
  ret
}

```

What this does for a single species:
```{r, get_presabs_summary_eg}
get_presabs_sp_summ(sp_index = spp$aphia_index[1])
```

Run across species and summarise (takes a minute or two to run):

```{r, get_presabs_sunmary_allspp}
pres_abs_summ <- spp %>% pull(aphia_index) %>%
  purrr::map(get_presabs_sp_summ, .progress = TRUE) %>%
  bind_rows()

pres_abs_summ %>% group_by(presabs) %>% summarise(n_cases = sum(n))
```

So there is a manageable number of presences (~3.3M) but the number of absences (~2bn) and NAs (~389M) will cause issues.

Given that NA indicates an incomplete survey of the benthic community, the decision taken for this product is not to consider sampling events that include NA for any species. First, add a new variable to `sample_events`, `includes_na`, setting this to be `FALSE` for all sampling events initially:

```{r, create_includes_na}
sample_events <- sample_events %>% mutate(includes_na = FALSE)
```

This function finds for a given species which (if any) sampling events include NA, and changes the `includes_na` flag to `TRUE` for any events that do:
```{r, get_na_events}
get_na_events <- function(sp_index, samps = sample_events){
  
  # get pres-abs data
  presabs <- ncvar_get(nc_fil, "Pres_abs",
                       start = c(1, sp_index), count = c(-1,1))
  # flag NAs in sample_events
  samps <- samps %>% mutate(includes_na = ifelse(is.na(presabs), TRUE, includes_na))
  samps
}

```

Then run this over all species, saving the results to a new object called `na_events` (takes a couple of minutes):

```{r, get_na_events_allspp}
na_events <- sample_events
for(i in spp$aphia_index){
  na_events <- get_na_events(sp_index = i, samps = na_events)
}

```

What are the consequences of this?
```{r, summarise_includes_na}
na_events %>% count(includes_na)
```

So getting rid of these would remove `r round(100*(sum(na_events$includes_na) / nrow(na_events)), 1)`% of all sample events. Are these biased in time?

```{r, echo = FALSE, message = FALSE}
ggplot(na_events) + geom_histogram(aes(x = year, fill = includes_na), alpha = 2/3)
```

Not obviously - although we may lose some of the much older events.

### Final assembly of `sample_events`

To finalise `sample_events`, re-create it from `na_events`, add an id variable from the rownumber, and deal with the dates in the future (here simply by flagging them in the `includes_na` variable so they can be filtered out easily in one step):

```{r, finalise_sample_events}

sample_events <- na_events %>% 
  mutate(sample_id = row_number()) %>% 
  dplyr::select(sample_id, everything()) %>% 
  mutate(includes_na = ifelse(date > lubridate::now(), TRUE, includes_na))

```

This can then be written to file `(write_csv(sample_events, file = here("data", "EMODnet-Biology-Benthos-European-Seas/sample_events.csv"))` allowing the final assembled version of `sample_events` to be read in quickly here:

```{r, read_sample_events}
sample_events <- read_csv(here("data",
                               "EMODnet-Biology-Benthos-European-Seas/sample_events.csv"))

```

### Assemble species presence x site data frame 

This assembles the long species presence x site df, for non-na sites only. It uses a revised version of the previous `get_presabs_sp()` function which excludes NA sites:

```{r, get_presabs_sp_na_rm}

get_presabs_sp_na_rm <- function(sp_index, sp_df = spp, samps = sample_events){
  
  # get pres-abs data
  presabs <- ncvar_get(nc_fil, "Pres_abs",
                       start = c(1, sp_index), count = c(-1,1))
  # assemble data frame
  ret <- tibble(aphia_id = spp$AphiaID[spp$aphia_index == sp_index],
                presabs = presabs)
  ret <- samps %>% bind_cols(ret)
  
  # filter to presences and non-NA sites
  ret <- ret %>% filter(presabs == 1 & includes_na == FALSE) %>% 
    dplyr::select(sample_id, aphia_id)
  
  ret
}

```

Run for one species - this needs the `spp` table of species:

```{r, read_species_list, message = FALSE}
taxo <- read_csv(here("data", "EMODnet-Biology-Benthos-European-Seas/benthos_taxo_rank.csv"))
spp <- taxo %>% 
  filter(rank == "Species") %>% 
  mutate(aphia_index = match(.$AphiaID, taxo$AphiaID))
```

Then run the presence-absence function for a single species:
```{r, get_presabs_sp_na_rm_eg}
get_presabs_sp_na_rm(sp_index = spp$aphia_index[1])
```

Now run over all species to generate a full dataframe of species presence at each site (excluding sites with NA values) - this takes around 5 minutes:

```{r, get_presabs_sp_na_rm_allspp}

pres_df <- spp %>% pull(aphia_index) %>%
  purrr::map(get_presabs_sp_na_rm, .progress = TRUE) %>%
  bind_rows()

```
From this we can get (and plot) e.g. a summary of species occurrence frequencies:

```{r, get_sp_occ_summary, message = FALSE}
sp_occ_summary <- pres_df %>% count(aphia_id) %>% arrange(n)
ggplot(sp_occ_summary) + geom_histogram(aes(x = n))
```
  
So for instance `r sum(sp_occ_summary$n == 1)` species occur in only a single sample; `r sum(sp_occ_summary$n > 1)` occur in more than one.

Before closing the nc connection, first extract the Coordinate Reference System:

```{r, get_presabs_crs}
presabs_crs <- ncatt_get(nc_fil, "crs")
```

Also worth saving the presence data (`pres_df`) to file for direct loading (`write_csv(pres_df, file = here("data", "EMODnet-Biology-Benthos-European-Seas/pres_df.csv"))`) allowing the final presence data frame to be read in quickly here:

```{r, read_pres_df}
pres_df <- read_csv(here("data",
                         "EMODnet-Biology-Benthos-European-Seas/pres_df.csv"))

```

Finally close the nc connection:
```{r, close_nc}
nc_close(nc_fil)
```
